Asteroseismology: AI-Driven Seismic Event Detection for Extra-Terrestrial Missions\\

### 2025-05-25 paper.tex
added conventinal content in methodalogy

\title{Asteroseismology: AI-Driven Seismic Event Detection for Extra-Terrestrial Missions\\
    % {\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and
    % should not be used}
    \thanks{This work was supported in part by [Identify applicable funding agency here if any, otherwise delete or state self-funded].}
}

@@

    \subsection{Initial Processing and Candidate Events Selection(IPCS)}%TODO refine heading
    (This entire subsection is a major differentiator from purely ML-driven approaches like Civilini et al.'s direct
    spectrogram-to-CNN method. You're showing a sophisticated "front-end".)

        \subsubsection{Adaptive Bandpass Filtering}
        (Your method for "Freq AnimationFinding the Best Bandpass Frequency": power spectrum analysis or standard
            deviation analysis. Tunable parameters involved.)

            \subsubsection{Noise and Outlier Mitigation}
            (Techniques for removing short, high-amplitude noises and other artifacts.)

        \subsubsection{Data Normalization}
            (Method used for consistent scaling.)

        \subsubsection{STA\slash LTA Analysis for Initial Detection}
            (Standard STA\slash LTA application: window lengths, thresholds. This is a common baseline.)

        \subsubsection{Candidate Pick Refinement using Characteristic Function Analysis}
        \label{sec:candidate_refinement}
            (Detail this specific filtering mechanism based on rise/fall slopes. This is likely a novel or specifically
            adapted part of your conventional pipeline and should be emphasized.)

@@

    \subsubsection{Input Feature Construction for the CNN}
        (How candidate events from \ref{sec:candidate_refinement} are prepared for the CNN.) 1D Seismogram Segments:
        Length (e.g., 5565 samples), channels used. Auxiliary Features: Explicitly detail the "Standard deviation values
        before and after the Arrival time." Justify why these features are chosen and what information they provide to
        the network. (Highlighting point: Civilini et al. used 2D spectrograms. Your use of 1D waveforms + explicitly
        engineered auxiliary features is a key methodological difference.)

@@

    % \bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350. 

>>>>

\title{Asteroseismology: AI-Driven Seismic Event Detection for Extra-Terrestrial Missions\\
}

@@

    \subsection{Initial Processing and Candidate Events Selection(IPCS)}%TODO refine heading
        The Initial Processing and Candidate Selection (IPCS) stage of the framework employs a series of conventional
        seismic signal processing techniques to enhance signal clarity, identify potential seismic events, and reduce
        the number of candidates passed to the subsequent machine learning stage. This stage operates on individual
        seismogram segments loaded from MiniSEED files.

        \subsubsection{Adaptive Bandpass Filtering}
            To isolate frequencies most likely to contain seismic signals and attenuate background noise, an adaptive
            bandpass filtering approach is implemented. The system iteratively evaluates a series of predefined
            frequency ranges. For each range, defined by a starting frequency  and a fixed window size , a bandpass
            filter is applied to a copy of the original trace.

            The optimal frequency band is determined by one of two methods, selectable by the user:

            \begin{enumerate}[label=\alph*), leftmargin=3em]
                \item \textit{Power Spectrum Analysis:} The spectrogram of the filtered signal is computed.
                The power values across all frequencies and time points within the spectrogram are flattened, and the average of
                a predefined number (e.g., top 50 or 1000) of the highest power values is calculated. The frequency range
                yielding the maximum average power among these top values is selected as the optimal band. This method
                prioritizes frequency bands where signal energy is most concentrated.
                \item \textit{Standard Deviation Analysis:} The filtered trace data is normalized, and its
                standard deviation is calculated. The frequency range that results in the minimum standard deviation of the
                normalized filtered signal is chosen as the optimal band. This method aims to find a frequency band where the
                signal is most distinct from broadband noise, assuming that a lower standard deviation in a filtered signal
                (after normalization) might indicate a clearer, less noisy signal.
            \end{enumerate}

        \subsubsection{Noise and Outlier Mitigation}
            Following the adaptive bandpass filtering, further steps are taken to mitigate noise. If resampling is
            configured , the filtered trace is resampled to a target frequency (e.g., 6.625Hz, reminiscent of Apollo-era
            sampling rates, though this is a tunable parameter). Subsequently, a threshold-based outlier removal is applied.
            Data points in the filtered trace whose absolute values exceed a multiple (e.g., 26 times) of the trace's
            standard deviation are considered outliers and are clipped to the standard deviation value. This step aims to
            reduce the impact of short, high-amplitude glitches that might otherwise distort normalization or trigger false
            STA/LTA detections.

        \subsubsection{Data Normalization}
            After outlier removal, the filtered trace data is normalized to a consistent amplitude range,
            typically between -1 and 1. This is achieved using min-max scaling:

            Normalization ensures that subsequent amplitude-based detection algorithms, like STA/LTA, operate on a
            consistent scale regardless of the original signal's absolute amplitude.

        \subsubsection{STA\slash LTA Analysis for Initial Detection}
            Potential seismic event onsets are identified using the classic Short-Term Average to Long-Term Average
            (STA/LTA) algorithm, as implemented in Obspy \cite{Beyreuther2010}. The algorithm is applied
            to the absolute values of the normalized and filtered trace data. The lengths of the short-term window
            and long-term window, as well as the trigger-on threshold and trigger-off threshold, are configurable parameters. The trigger-off threshold is dynamically set based on the average of the
            characteristic function (CFT) generated by the STA/LTA algorithm. This results in a set of on-off time pairs,
            marking the start and end times of candidate detections.

        % \subsubsection{Candidate Pick Refinement using Characteristic Function Analysis}
        %     Not all STA/LTA triggers correspond to genuine seismic events. The Asteroseismology framework incorporates a
        %     refinement step based on the duration and number of STA/LTA triggers. Specifically, from the on-off pairs
        %     generated by STA/LTA, only those triggers where the duration (triggers[1] - triggers[0]) exceeds a certain
        %     minimum (e.g., 1200 samples, equivalent to approximately 3 minutes at 6.625 Hz if resampled) OR if the total
        %     number of triggers for the seismogram is below a certain count (e.g., less than 10) are considered for
        %     further processing. This heuristic aims to filter out very short, isolated triggers that are less likely to
        %     be significant seismic events while retaining detections from seismograms with sparse activity. Events
        %     passing this refinement are then considered qualified candidates for the subsequent machine learning stage.
        %     [Note: The provided code's characteristic function slope analysis is not explicitly implemented in the
        %     detect_events function snippet, but rather implied as a general concept in your abstract. If you have a
        %     separate code module for that, we'd need to describe it. The current code filters based on duration/count of
        %     STA/LTA triggers.]
        % \label{sec:candidate_refinement}
        %     (Detail this specific filtering mechanism based on rise/fall slopes. This is likely a novel or specifically
        %     adapted part of your conventional pipeline and should be emphasized.)

@@

    \subsubsection{Input Feature Construction for the CNN}
        (How candidate events from are prepared for the CNN.) 1D Seismogram Segments:
        Length (e.g., 5565 samples), channels used. Auxiliary Features: Explicitly detail the "Standard deviation values
        before and after the Arrival time." Justify why these features are chosen and what information they provide to
        the network. (Highlighting point: Civilini et al. used 2D spectrograms. Your use of 1D waveforms + explicitly
        engineered auxiliary features is a key methodological difference.)

@@

    \bibitem{Beyreuther2010} M. Beyreuther, R. Barsch, L. Krischer, T. Megies, Y. Behr, and J. Wassermann, "ObsPy: A Python toolbox for seismology," Seismological Research Letters, vol. 81, no. 3, pp. 530-533, 2010.   % \bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350. 
